---
title: "Analysis- Personal Activity Monitoring"
author: "Urvesh ******"
date: "Sunday, August 16, 2015"
output: html_document
---
```{r global_options, include = FALSE}
knitr::opts_chunk$set(warning = FALSE)
```

This is a report on analysis performed on the Personal Activity Monitoring Data.
Data is available at [Personal Activity data](https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip)

============================================================================  
__Loading and preprocessing the data__

```{r}
data <- read.csv("activity.csv")
summary(data)
```

============================================================================  
__What is mean total number of steps taken per day?__

1. Calculating the number of steps per day using aggregate function
```{r}
dataagg <- aggregate(data$steps ~ data$date, data, FUN = "sum")
names(dataagg)<- c("day","steps")
head(dataagg)
```

2. Let us plot the histogram of steps and see how it looks like.
We are also calculating mean and median values.
```{r hist_total_no_of_step_taken_each_day}
hist(dataagg$steps, xlab = "steps per day", main = "Histogram of steps per day", col ="gray")
```

3. Calculating mean and median
```{r}
meanPerday <- as.integer(mean(dataagg$steps))
medianPerday <- median(dataagg$steps)
meanPerday
medianPerday
```

Mean value for the steps per day is __`r meanPerday`__  
Median value for the steps per day is __`r medianPerday`__

============================================================================  
__What is the average daily activity pattern?__

1. creating time series plot of the 5-min interval and the average number of  
steps taken, averaged across all days
```{r}
dataagg1 <- aggregate(data$steps ~ data$interval, data, FUN = "mean")
names(dataagg1) <- c("interval","stepsAvg")
```

```{r plot_avg_no_of_steps_taken_vs_interval}
plot(dataagg1$interval, dataagg1$stepsAvg, type = "l", xlab = "interval",
     ylab = "Average steps")
```

2. Finding interval which contains maximum number of averaged steps
```{r}
maxSteps <- max(dataagg1$stepsAvg)
maxInt <- dataagg1[dataagg1$steps==maxSteps,1]
maxInt
```
The interval which has maximum averaged steps is __`r maxInt`__

============================================================================  
__Imputing missing values__

1. Calculating total number of missing values in dataset

```{r}
sum(is.na(data$steps))
```

2. imputing the missing values.  
__Strategy__ here is to fill missing value with 
average number of steps for that interval across all days  
(note: we are using previously calculated avg steps for intervals across days)
3. completeData is the new dataset with missing data filled in
```{r}
completeData <- data
for (i in 1:nrow(completeData))
  if(is.na(completeData[i,1]))
    completeData[i,1] <- dataagg1[dataagg1$interval==completeData[i,3],2]  
```

4. Let us calculate new histogram, mean and meadian values per day
```{r}
dataaggNew <- aggregate(completeData$steps ~ completeData$date, completeData, FUN = "sum")
names(dataaggNew)<- c("day","steps")
meanPerdayNew <- as.integer(mean(dataaggNew$steps))
medianPerdayNew <- as.integer(median(dataaggNew$steps)) 
```

```{r hist_total_no_of_step_taken_each_day_after_imputing}
hist(dataaggNew$steps, xlab = "steps per day", main = "new Histogram of steps per day", col="gray")
```

mean value before: __`r meanPerday`__ mean value after: __`r meanPerdayNew`__  
median value before: __`r medianPerday`__ median value after: __`r medianPerdayNew`__

So It is clear that there is no major difference in mean and median values

============================================================================  
__Are there differences in activity patterns between weekdays and weekends?__

1. creating new factor variable with factors 2 weekday and weekend

```{r}
completeData$day <- weekdays(as.POSIXct(completeData$date))

for (i in 1:nrow(completeData))
  completeData[i,4] <- if (completeData[i,4] == "Sunday" || completeData[i,4] == "Saturday") "weekend" else "weekday" 

completeData$day  <- as.factor(completeData$day)
```

2. Creating panel plot of the 5-minute interval and the average number of steps taken,  
averaged across all weekday days or weekend days (y-axis)

first let us calculate average number of steps avarage weekdays/weekend
```{r}
dataaggNew1 <- aggregate(completeData$steps ~ completeData$interval + completeData$day, data, FUN = "mean")
names(dataaggNew1) <- c("interval","day","stepsAvg")
```

Time to plot...  
```{r Panelplot}
require("lattice")
xyplot(stepsAvg ~ interval | day, dataaggNew1, type = "l", layout = c(1, 2), xlab = "Interval",
       ylab = "Number of steps")
```

Looking at the plot, it is clear that there is a difference in activity patterns on weekdays and weekends. We can observe high number of states on weekend for interval values ranging from 1000 to 2000 
